---
title: "ENVS193DS-homework-03"
author: "Kate Bierwagen"
date: "05-28-2026"
format: pdf
---
# Setting Up
```{r}
#| message: false

# loading in packages 
library(tidyverse)
library(here)
library(janitor)
library(readxl)
```

```{r}
#| message: false
# reading in data
kelp <- read_csv(here("data", "temp-kelp.csv"))
```

# Problem 1. Giant kelp fronds

## a. An appropriate test

Since you are interested in the strength of the relationship between temperature and giant kelp frond elongation rate, the appropriate test is a Pearson's r or a Spearman rank correlation, which are both tests for correlation. Pearson's r test is a parametric version of a correlation test which is used for continuous and normally distributed variables, testing the linear relationship between them. Spearman rank correlation is a non-parametric version, which is testing the monotonic relationship between variables.

## b. Create a visualization!!!










## c. Check your assumptions and run your test!!




## d. Results communication



## e. Test implications



## f. Double check your own work.



